What We Eval in the Shadows

### 2021-04-04 paper/main.tex
Fix layout in S4

\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
as a few other ones that allow us to better identify where do the expressions
come from. For example, we taint the results of calls to \c{parse} (turns
string into expression) and \c{match.call} (reflects on current call). We
also capture calls to R API for dynamic code loading (\eg\xspace \c{library},
\c{require} and \c{source}). Next, we subscribe to the events related to a
variable definition and assignment allowing us to record side effects that
happen in environments while evaluating code in \eval. One challenge we
encountered during the analysis was that essentially R only provides source
references for top-level calls in a function body or in the block surrounded by
braces. That is to say, a method whose body would \c{eval(x)} would not have
debug information that we can use to identify that particular \eval. The same
expression surrounded by braces would have a line number and source reference.
This is unfortunately not easily fixed in the R implementation. Instead, we
extended the dynamic analysis tool to attach synthetic source code references
to all \eval call sites by traversing AST.  The tracer is implemented as an R
package in 3.2K C++ and 1.3K of R code. For performance reasons, most of the
tracing is done in C++. While in theory the implementation is rather
straightforward, what makes it hard in practice is that (1) some of the core R
functionality such as package loading is implemented using \eval which then
leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
to analyze function arguments while tracing (3) running large amount of
real-world code triggers exercises a lot of corner cases of highly
underspecified R behavior.

@@

@@

\begin{figure}[!h]

@@

\begin{wrapfigure}{r}{5.8cm}
\vspace{-4mm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}

>>>>

\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
as a few other ones that allow us to better identify where do the expressions
come from. For example, we taint the results of calls to \c{parse} (turns
string into expression) and \c{match.call} (reflects on current call). We also
capture calls to R API for dynamic code loading (\eg\xspace \c{library},
\c{require} and \c{source}). Next, we subscribe to the events related to a
variable definition and assignment allowing us to record side effects that
happen in environments while evaluating code in \eval. One challenge we
encountered during the analysis was that essentially R only provides source
references for top-level calls in a function body or in the block surrounded
by braces. That is to say, a method whose body would \c{eval(x)} would not have
debug information that we can use to identify that particular \eval. The same
expression surrounded by braces would have a line number and source reference.
This is unfortunately not easily fixed in the R implementation. Instead, we
extended the dynamic analysis tool to attach synthetic source code references
to all \eval call sites by traversing AST.  The tracer is implemented as an R
package in 3.2K C++ and 1.3K of R code. For performance reason, most of the
tracing is done in C++. While in theory the implementation is rather
straightforward, what makes it hard in practice is that (1) some of the core R
functionality such as package loading is implemented using \eval which then
leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
to analyze function arguments while tracing as prematurely forcing a promise
might have consequences, and finally (3) running large amount of real-world
code triggers exercises a lot of corner cases of highly underspecified R
behavior.

@@

\newpage

@@

\begin{figure}[b]

@@

\begin{wrapfigure}{4}{5.8cm}
\vspace{-4mm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
